---
title: "Repdata-PA N1"
author: "Alessio Ballesio"
date: "27/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setting up analysis by set the working directory and read the data

```{r}
setwd("C:/Users/aless/Documents/data science/course n5 - reproducible research/Repdata-peerassessment1")
activity <- read.csv("activity.csv", header= TRUE)
```

First question asks to calculate the mean of steps. To do this I first need to calculate the number of steps done each day. 
Then I first plot histogram of the results using the base plot systems
As second step, calculate the mean and median values
```{r}
total_daily_steps<- tapply(activity$steps,activity$date,sum,na.rm = TRUE)
histogram <- hist(total_daily_steps, xlab= "Number of daily steps", main = "Distribution of daily steps")
dev.copy(png, file = "Distribution of daily steps.png") 
dev.off()
```
```{r}
mean_daily_steps <- mean(total_daily_steps)
mean_daily_steps
median_daily_steps <- median(total_daily_steps)
median_daily_steps
```

## SECONd QUESTION - What is the average daily activity pattern?

First of all I need to calculate, using again tapply function the average steps for each interval across all days. 
In order to make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days (y-axis). I then plot the function.

```{r average daily activity, echo=TRUE}
library(dplyr)
intervals <- group_by(activity, interval)
average_steps_interval <- summarise(intervals,steps=mean(steps,na.rm=TRUE))
# putting into a new png file
plot(average_steps_interval, type= "l", col = "red", xlab = "Interval", ylab= "average n. of steps")
dev.copy(png, file = "average daily activity.png")
dev.off()
```

The second part of question asks to determine which is interval with the highest average number of steps. For this, I use the which.max function
```{r , echo=TRUE}
max_interval <- average_steps_interval[which.max(average_steps_interval$steps),]
max_interval
```

THIRD STEP - Imputing missing values

First of all I calculate the total number of rows (intervals) for which there is no value. I use function sum(is.na())

```{r , echo=TRUE}
Number_missing_values <- sum(is.na(activity$steps))
Number_missing_values
```

I then decide to replace each NA, with the average value across all intervals- I first create a logical vector which tells in which rows there is a missing value. Then I calculate the mean across all intervals.
I create therefore a processed "activity" dataset, where instead of those missing values, I put the mean across all intervals, using the logical vector missing_values
```{r replacing missing value, echo=TRUE}
missing_values <- is.na(activity$steps)
value_to_add <- mean(activity$steps,na.rm=TRUE)
activity_processed <- activity
activity_processed[missing_values,3] <- value_to_add
```
I then create the histogram of this new dataset 
```{r}
total_daily_steps_processed<- tapply(activity_processed$steps,activity$date,sum,na.rm = TRUE)
histogram <- hist(total_daily_steps_processed, xlab= "Number of daily steps", main = "Distribution of daily steps")
dev.copy(png, file = "Distribution of daily steps-with NA replaced.png") 
dev.off()
```
and calcuate the mean and median
```{r}
mean_daily_steps_processed <- mean(total_daily_steps_processed)
mean_daily_steps_processed
median_daily_steps_processed <- median(total_daily_steps_processed)
median_daily_steps_processed
```

PART 4 - Are there differences in activity patterns between weekdays and weekends?

First step is to create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
activity_processed$day <- weekdays(as.Date(activity_processed$date))
activity_processed$day_type <- ifelse(activity_processed$day=='sabato' | activity_processed$day=='domenica', 'weekend','weekday')
```

Then I pass to create a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
average_steps_interval_and_daytype <- aggregate(steps~interval+day_type,data=activity_processed,FUN=mean,na.action=na.omit)
library(ggplot2)
plot <- ggplot(average_steps_interval_and_daytype,aes(interval,steps))
plot+geom_line(col="red") + facet_grid(day_type ~ .)+ xlab("interval")+ylab("average n. of steps")
dev.copy(png, file= "weekdays vs weekend.png")
dev.off()
```

END